Research Experience

### 2010-04-06 doc/thesis/thesis.tex
quick addition to thesis. fix to svn version tagging.
latex runs correctly now.

\subsection{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_poststim}, though the reason for this
is debated \cite{origin_poststim}.

@@

\subsection{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{ModelCompareison}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

@@

For \{delta, DC/Spline\}, \{exponential, gaussian, cauchy\}, \{biased, unbiased initial\},
\{100, 500, 1000\} particles
\begin{enumerate}
\item Comparison with a linear system with similar number of degrees of freedom
\item Ground truth vs. Estimated signal during particle filter run
\item Ground truth vs. Estimated signal with final parameter set
\item True Parameters vs. Final Parameter Sets
\item Variance of final parameters when faced with same ground truth, different noise
\item Variance of final parameters when faced with same ground truth, different noise
\item MSE of (a new timeseries based on X(t) vs. ground truth) for all t
\item Estimator Variance based on different noise runs
\end{enumerate}

>>>>

\subsection{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_undershoot}, though the reason for this
is debated \cite{origin_poststim}.

@@

\subsection{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{nonlinearmodels}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

@@

For \{delta, DC/Spline\}, \{exponential, gaussian, cauchy\}, \{biased, unbiased initial\},
\{100, 500, 1000\} particles
\begin{enumerate}
\item Comparison with a linear system with similar number of degrees of freedom
\item Ground truth vs. Estimated signal during particle filter run
\item Ground truth vs. Estimated signal with final parameter set
\item True Parameters vs. Final Parameter Sets
\item Variance of final parameters when faced with same ground truth, different noise
\item Variance of final parameters when faced with same ground truth, different noise
\item MSE of (a new timeseries based on X(t) vs. ground truth) for all t
\item Estimator Variance based on different noise runs
\item Final Particle Distribution
\end{enumerate}
